---
title: "WGCNA"a
author: "Lin Hou"
date: "2024-06-12"
output: html_document
source: https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0
The WGCNA R package builds “weighted gene correlation networks for analysis” from expression data. It was originally published in 2008 and cited as the following:

Langfelder, P. and Horvath, S., 2008. WGCNA: an R package for weighted correlation network analysis. BMC bioinformatics, 9(1), p.559.
Zhang, B. and Horvath, S., 2005. A general framework for weighted gene co-expression network analysis. Statistical applications in genetics and molecular biology, 4(1).
---

```{r setup, include=FALSE}
library(WGCNA)
library(ggplot2)
library(tidyverse)
library(magrittr) 
library(dplyr)
library("igraph")
knitr::opts_chunk$set(warning = FALSE)
```


```{r import dataset}
setwd("/Users/houlin/Desktop/Research/Ecohab/Ecohab22/metatranscriptoms")
df_org<-read.csv('CPL/CPL_filtered_tax_full_altered_MCL_cluster.csv',row.names="geneID")

auto_list=c('Archaeplastida','Synurophyceae','Cryptophyta','Haptophyta','Filosa-Chlorarachnea','Dinophyceae','Filosa-Chlorarachnea','Bacillariophyta','Bolidophyceae-and-relatives','Chrysophyceae-Synurophyceae','Dictyochophyceae','Pelagophyceae','Phaeophyceae','Pinguiophyceae','Raphidophyceae','Synchromophyceae','Xanthophyceae')
autopattern <- paste(auto_list, collapse = "|")

metadata<-read.table('/Users/houlin/Desktop/Research/Ecohab/Ecohab22/data/Ecohab22_metadata_samples.tsv', header = TRUE, sep = "\t")
metadata$Starttime <- as.POSIXct(metadata$Starttime)
# Extract month and day formatted as "MM-DD"
metadata$month_day <- format(metadata$Starttime, "%m-%d")
rownames(metadata) <- metadata$`sample.id`
metadata$`sample.id` <- NULL
```

```{r}
tax<-'Autotrophs'

if (tax == "noMET") {
    df_use <- df_org[df_org$Kingdom != "Opisthokonta", ]
  } else if (tax == "Autotrophs") {
    df_use <- df_org[grepl(autopattern, df_org$LPI_taxonomy), ]
  } else {
    df_use <- df_org[grepl(tax, df_org$LPI_taxonomy), ]
  }

df_cluster<- df_use%>%
  group_by(MCL_cluster) %>%
  summarise(across(0:63, sum, na.rm = TRUE)) %>%
  as.data.frame()
rownames(df_cluster) <-  df_cluster$MCL_cluster

df_cluster <- df_cluster %>%
  select(-MCL_cluster)
df_cluster<-df_cluster[-1,]
print(paste(tax,'MCL_clustered dataframe created!'))
```
```{r}
big_cluster<-df_use[df_use$MCL_cluster_size>10,'MCL_cluster']
big_cluster <- unique(big_cluster)
big_cluster<-big_cluster[-1]
big_cluster <- union(big_cluster, c("dabA", "dabC", "dabD"))
```


```{r}
df_relative <- df_cluster / colSums(df_cluster)*100
mean_expression <- rowMeans(df_relative)
# Identify genes with mean expression > 0.1% (0.01 and 0.05 for diatom only)
# 0.1% and 0.5% --> 22433,0.01% and 0.05%--> 106474
# 0.25% B. C. Kolody et al. and 0.5% ---> 11465 #11168
thres<-0.005
condition_mean <- mean_expression > thres
# Identify genes with any sample expression > 0.5%
condition_sample <- apply(df_relative, 1, function(x) any(x > 0.1))
keep_clusters <- rownames(df_relative) %in% c("dabA", "dabC","dabD")
genes_to_keep <- condition_mean | keep_clusters #| condition_sample
genes_to_keep[is.na(genes_to_keep)] <- FALSE
#filtering for the datasets
# keep only rownames that are in big_cluster as well
#df_filtered_cluster <- df_cluster[genes_to_keep & rownames(df_cluster) %in% big_cluster, ]
#df_filtered_relative <- df_relative[genes_to_keep & rownames(df_relative) %in% big_cluster, ]

df_filtered_cluster <- df_cluster[genes_to_keep, ]
df_filtered_relative <- df_relative[genes_to_keep, ]
### filtering for counts > 0 for over 50% of the samples
#keep <- apply(df_filtered_cluster, 1, function(x) sum(x > 0) > (0.25 * ncol(df_filtered_cluster)))
#df_filtered_cluster<-df_filtered_cluster[keep,]
#df_filtered_relative <- df_filtered_relative[keep,]

datExpr <- log2(df_filtered_cluster + 1)

datExpr_relative <- log2(df_filtered_relative + 1)

print(dim(datExpr))
```


```{r Creating expression plots}
df<-datExpr #[turquoise_genes,]

expr_normalized_df <- data.frame(df) %>%
  mutate(Gene_id = row.names(df)) %>%
  pivot_longer(-Gene_id) # transform the data from wide into long format

expr_normalized_df %>% ggplot(., aes(x = name, y = value)) +
  geom_violin() +geom_point() +theme_bw() +
  theme(axis.text.x = element_text( angle = 90)) +
  ylim(0, NA) +
  labs(title = paste0(tax," Normalized  Expression"),
    x = "Samples",
    y = "normalized expression")
```

```{r WCGNA transposed dataset}
df<-datExpr
input_mat <- t(df) # transposed dataset
allowWGCNAThreads(nThreads = 8)        
powers = c(c(1:10), seq(from = 12, to = 20, by = 2))
sft = pickSoftThreshold(
  input_mat,             # <= Input data
  #blockSize = 30,
  powerVector = powers,
  verbose = 1)

par(mfrow = c(1,2));
cex1 = 0.9;
plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2",
     main = paste(tax,"Scale independence"))

text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red")
abline(h = 0.90, col = "red")
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste(tax,"Mean connectivity"))
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers,
     cex = cex1, col = "red")
```

Pick a soft threshold power near the curve of the plot, so here we could pick 7, 8 or 9. We’ll pick 9 but feel free to experiment with other powers to see how it affects your results. Now we can create the network using the blockwiseModules command. The blockwiseModule may take a while to run, since it is constructing the TOM (topological overlap matrix) and several other steps. Parameters: https://www.rdocumentation.org/packages/WGCNA/versions/1.69/topics/blockwiseModules

```{r WCGNA network}
start_time <- Sys.time()
picked_power = 10
temp_cor <- cor       
cor <- WGCNA::cor         # Force it to use WGCNA cor function (fix a namespace conflict issue)
netwk <- blockwiseModules(input_mat,                # <= input here
                          # == Adjacency Function ==
                          power = picked_power,                # <= power here
                          networkType = "signed",
                          corFnc = "cor",                     
                          corOptions = list(method = "spearman"),
    
                          # == Tree and Block Options ==
                          deepSplit = 2,
                          pamRespectsDendro = F,
                          # detectCutHeight = 0.75,
                          minModuleSize = 50,
                          maxBlockSize = 10000,
                          # == Module Adjustments ==
                          reassignThreshold = 0.25,
                          mergeCutHeight = 0.25,
                          # == TOM == Archive the run results in TOM file (saves time)
                          saveTOMs = T,
                          saveTOMFileBase = "ER",
                          # == Output Options
                          numericLabels = T,
                          verbose = 3)
cor <- temp_cor
end_time <- Sys.time()
runtime <- end_time - start_time
print(runtime)
```

Let’s take a look at the modules, there
```{r WCGNA plotting}
# Convert labels to colors for plotting

mergedColors <- labels2colors(netwk$colors)
ModuleColors <- mergedColors[netwk$blockGenes[[1]]]
geneTree<-netwk$dendrograms[[1]]
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(
  geneTree,
  ModuleColors,
  "Module colors",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05,
  main = paste(tax, "CPL Gene dendrogram >", thres,'%'))
```


Relate Module (cluster) Assignments to Treatment Groups

WGCNA will calcuate an Eigangene (hypothetical central gene) for each module, so it easier to determine if modules are associated with different samples.

```{r Visualizing Module's Association with different samples}
# Get Module Eigengenes per cluster
MElist <- moduleEigengenes(input_mat, mergedColors)
MEs<-MElist$eigengenes 
MEs0 <- MElist$eigengenes 

# Reorder modules so similar modules are next to each other
MEs0 <- orderMEs(MEs0)
module_order = names(MEs0) %>% gsub("ME","", .)

# Add samples names
#MEs0$samples = row.names(MEs0)
MEs0$samples= metadata$Phase_group
# tidy & plot data
mME = MEs0 %>%
  pivot_longer(-samples) %>%
  mutate(
    name = gsub("ME", "", name),
    name = factor(name, levels = module_order))
```

```{r}
ME.dissimilarity = 1-cor(MElist$eigengenes, use="complete")
METree = hclust(as.dist(ME.dissimilarity), method = "average") #Clustering eigengenes 
par(mar = c(0,4,2,0)) #seting margin sizes
par(cex = 0.6);#scaling the graphic
plot(METree)
abline(h=.3, col = "red") #a height of .25 corresponds to correlation of .75

merge <- mergeCloseModules(input_mat, mergedColors, cutHeight = .3)
New_mergedColors = merge$colors
New_ModuleColors<-New_mergedColors[netwk$blockGenes[[1]]]
New_mergedMEs = merge$newMEs

plotDendroAndColors(geneTree, cbind(ModuleColors, New_ModuleColors), 
c("Original Module", "Merged Module"),
dendroLabels = FALSE, hang = 0.03,
addGuide = TRUE, guideHang = 0.05,
main = paste(tax,"CPL Gene dendrogram >", thres,"%"))

```


```{r}
metadata_df<-metadata[,c("DA","Mean_Temp","Mean_Salinity","Mean_Chl","shannon_entropy_prokaryotes",
                         "shannon_entropy_eukaryotes","shannon_entropy_auto")]
Samples <- rownames(input_mat)
sampleRows <- match(Samples, rownames(metadata_df))
metadata_df<- metadata_df[sampleRows,]
```


```{r}
nGenes = ncol(input_mat)
nSamples = nrow(input_mat)
MEs<-orderMEs(MEs)
module.trait.correlation = cor(MEs, metadata_df,use = 'p', method = 'spearman') #p for pearson correlation coefficient 
module.trait.Pvalue = corPvalueStudent(module.trait.correlation, nSamples)
module=names(MEs) %>% gsub("ME","", .)
textMatrix = paste(signif(module.trait.correlation, 2), "\n(",signif(module.trait.Pvalue, 1), ")", sep = "")
dim(textMatrix) = dim(module.trait.correlation)
mat<-module.trait.correlation

png(filename = paste0("/Users/houlin/Desktop/Research/Ecohab/Ecohab22/metatranscriptoms/WGCNA/Auto_005_Jan4_run/",tax,"_",thres,"_Module-Trait_Relationship.png"),
    width = 20, height = 18, res=300, unit="cm")
annotationMatrix <- ifelse(module.trait.Pvalue < 0.05, "*", "")
par(mar = c(9, 7, 2, 2))
labeledHeatmap(Matrix = mat,
xLabels = names(metadata_df),
yLabels = names(MEs),
ySymbols = module,
colorLabels = FALSE,colors = blueWhiteRed(50),
textMatrix = annotationMatrix,setStdMargins = FALSE,
cex.text = 1.5,zlim = c(-1,1),
main = paste(tax,"MCL_cluster Module-trait relationships CPL", thres,"%"))
```


```{r}
### renaming the parameters ######

salinity = as.data.frame(metadata_df$Mean_Salinity);
names(salinity) = "salinity"
chl = as.data.frame(metadata_df$Mean_Chl);
names(chl) = "chloropyhll"
temp = as.data.frame(metadata_df$Mean_Temp);
names(temp) = "temperature"

### Plotting the heatmaps ######
MEs<-orderMEs(cbind(MEs,salinity,chl,temp))
par(cex = 0.9)
plotEigengeneNetworks(MEs, "", marDendro = c(0,4,1,2), marHeatmap = c(5,6,1,2), cex.lab = 0.8, xLabelsAngle
= 90)

par(cex = 1.0)
plotEigengeneNetworks(MEs,paste(tax,"Eigengene Dendrogram"), marDendro = c(0,4,2,0),
plotHeatmaps = FALSE)

par(cex = 1.0, mar = c(4,4,1,1))
plotEigengeneNetworks(MEs, paste(tax,"Eigengene adjacency heatmap"), marHeatmap = c(7,7,2,2),
plotDendrograms = FALSE, xLabelsAngle = 90)
```
identify hub genes, i.e genes with the highest connectivity to each modules
```{r}
topHubs <- chooseTopHubInEachModule(input_mat, ModuleColors, power = picked_power,type = "signed")

adjacencyMatrix <- adjacency(input_mat, power = picked_power, type = "signed")
IMC <- intramodularConnectivity(adjacencyMatrix, mergedColors)

geneInfo <- data.frame(
    Gene = colnames(input_mat),
    Module = mergedColors,
    Connectivity = IMC$kWithin
)
rankedGenes <- geneInfo %>%
  arrange(Module, desc(Connectivity)) %>%
  group_by(Module) %>%
  mutate(Rank = row_number())
topGenes <- rankedGenes %>% filter(Rank <= 20) 

### Module membership, identifying correlation of genes to the modules
kME_table <- signedKME(input_mat,
                       MEs,
                       corFnc = "cor", 
                       corOptions = "use = 'p'")#, method = 'spearman'")

# Extract kME for each gene relative to its assigned module
kME_assigned <- sapply(1:ncol(input_mat), function(i) {
  module <- mergedColors[i]
  kME_table[i, paste0("kME", module)]  # Extract kME for the gene's assigned module
})

# Create a data frame with kME and module information
kME_results <- data.frame(
  Gene = colnames(input_mat),
  Module = mergedColors,
  kME = kME_assigned
)
head(kME_results)

#Kleinberg hub and authority score for a network
#in TOM, the adjacency matrix is symetrical, thus hub and authority scores are the same, only in directed network, the scores are #different

adjacency <- adjacency(input_mat, power = picked_power,type = "signed")
TOM <- TOMsimilarity(adjacency, TOMType = "signed")

rownames(TOM) <- colnames(input_mat)
colnames(TOM) <- colnames(input_mat)
network <- graph.adjacency(TOM, mode = "directed", weighted = TRUE, diag = FALSE)

# Add signed weights
E(network)$weight <- E(network)$weight * sign(E(network)$weight)
# Add module color information
V(network)$ModuleColor <- mergedColors[V(network)$name]

hub_scores <- hub_score(network, weights = E(network)$weight)$vector
#authority_scores <- authority_score(network, weights = E(network)$weight)$vector

# Combine scores with gene and module color information
clust_scores <- data.frame(
  cluster = names(hub_scores),
  HubScore = hub_scores,
  #AuthorityScore = authority_scores,
  ModuleColor = mergedColors
)
# write.csv(clust_scores, "/Users/houlin/Desktop/Ecohab/Ecohab22/metatranscriptoms/WGCNA/hubscores_gene_025.csv", row.names = FALSE)

# Top hubs for each module
#top_hubs <- clust_scores %>%
#  group_by(ModuleColor) %>%
#  arrange(desc(HubScore)) %>%
#  slice_head(n = 5)

```


```{r}
edge_list <- data.frame(
  Source = rep(colnames(TOM), each = ncol(TOM)),  # Use the column names as source
  Target = rep(colnames(TOM), times = ncol(TOM)),  # Use the column names as target
  weight = as.vector(TOM)  # Adjacency values (e.g., TOM)
)

# Remove self-loops (edges where a node is connected to itself)
edge_list <- edge_list[edge_list$Source != edge_list$Target, ]
edge_list_filtered<-edge_list[edge_list$weight>0.5,]

savepath<-"/Users/houlin/Desktop/Research/Ecohab/Ecohab22/metatranscriptoms/WGCNA/"
write.csv(edge_list, paste0(savepath,tax,"_network_edges_",thres,".csv"), row.names = FALSE)
write.csv(edge_list_filtered, paste0(savepath,tax,"_network_edges_filtered_",thres,".csv"), row.names = FALSE)
```

```{r}
#pull out the list of the modules
today <- Sys.Date()
module_df <- data.frame(
  gene_id = names(netwk$colors),
  Modules = labels2colors(netwk$colors),
  new_colors= New_mergedColors
  #Connectivity = IMC$kWithin,
  #kME = kME_assigned,
  #hubscore=hub_scores
)  
# save the results as csv for further analysis
savepath<-"/Users/houlin/Desktop/Research/Ecohab/Ecohab22/metatranscriptoms/WGCNA/"
write.csv(module_df,paste0(savepath,tax,"_MCL_cluster_modules_",thres,"_",today,".csv"),row.names = FALSE) 
```

#visualize the network produced by the WGCNA TOM
```{r}
TOMp<-TOM
TOMp[TOMp<0.6]=0
network_graph <- graph.adjacency(TOMp, mode = "undirected", weighted = TRUE, diag = FALSE)
V(network_graph)$color <- as.factor(mergedColors)
V(network_graph)$size <- degree(network_graph, mode = "all")
plot(network_graph, 
     vertex.size = V(network_graph)$size,  # Node size by degree
     vertex.color = V(network_graph)$color, # Node color by module
     vertex.label.cex = 0.6,               # Adjust label size
     edge.width = 0.5,                    # Edge width
     main = "WGCNA Network - Genes by Module Colors")
```


