---
title: "Differential_gene_expression"
output: html_document
date: "2024-09-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(DESeq2)
library(tidyverse)
library(Matrix)
library(igraph)
library(magrittr) 
library(dplyr)
library(otuSummary)
library(reshape2) # has the melt function, which I use to wrangle data
library("BiocParallel")
pass <- function(x){x}
```

## R Markdown
DESeq 
normalization method: Size Factor Calculation (Median-of-Ratios Method)
For each gene, DESeq2 calculates the geometric mean of its counts across all samples.
For each sample, the count for each gene is divided by this gene’s geometric mean to get a ratio.
DESeq2 then computes the median of these ratios for each sample, which serves as the size factor for that sample.
This size factor scales the counts for each sample, allowing for effective normalization of sequencing depth and library size differences.

```{r Import all datasets}
setwd("/Users/houlin/Desktop/Research/Ecohab/Ecohab22/metatranscriptoms")
genecounts<- read.csv("genecounts/Ecohab22_genecounts_annotation.csv",row.names=1)
#cpl.data <- read.csv("CPL/EcoHAB22-CPL-norm-clustered-filtered-taxfull.csv",row.names="geneID")
#auto.cpl<- read.csv("df_autotrophs_cpl.csv",row.names="geneID")
metadata <- read.csv("/Users/houlin/Desktop/Research/Ecohab/Ecohab22/data/Ecohab22_metadata_042125.csv",row.names = 'sample.id')
rows_to_drop <- c("EH16_0070", "EH15_0064")
# Drop rows with specified row names
metadata$Phase_group <- as.factor(metadata$Phase_group)
cluster_annotate<-read.csv("/Users/houlin/Desktop/Research/Ecohab/Ecohab22/metatranscriptoms/CPL/cluster_annotation_df.csv",row.names="cluster")
```

```{r}
tpm<- read.csv("/Users/houlin/Desktop/Research/Ecohab/Ecohab22/metatranscriptoms/Ecohab22_TPM_annotated-taxfull.csv",row.names=1)
annotation<-tpm[97:119]
```

```{r}
silA_gene<-read.csv("/Users/houlin/Desktop/Research/Ecohab/Ecohab22/metatranscriptoms/SiLA_gene.csv",row.names=1)
annotation<- merge(annotation, silA_gene, by = "row.names", all.x = TRUE)
rownames(annotation) <- annotation$Row.names
annotation$Row.names <- NULL
```



```{r pressure, echo=FALSE}
#data<-genecounts[4:66]
diatom_counts <- genecounts[grepl("Bacillariophyta", genecounts$LPI_taxonomy), 4:66]
data<-diatom_counts
row_names <- rownames(data)
metadata <- metadata[(rownames(metadata) %in% colnames(data)), ]
data<-data[,rownames(metadata)]
#rownames(data) <- row_names
metadata$Diatom_clusters <- factor(
  metadata$Diatom_clusters,
  levels = c(0, 1, 2),
  labels = c("nonbloom", "bloom-1", "bloom-2")
)
dds <- DESeqDataSetFromMatrix(countData = data,
                              colData = metadata,
                              design = ~ Diatom_clusters)
# filter by keeping genes with sum above 10 counts
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]
dds<- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE)
output_dir <- "/Users/houlin/Desktop/Research/Ecohab/Ecohab22/metatranscriptoms/DESeq2_results/"
write.csv(as.data.frame(normalized_counts), file = paste0(output_dir,"diatom_normalized_counts.csv"))  

register(MulticoreParam(12))
start_time <- Sys.time()
dds.results <- DESeq(dds)
saveRDS(dds.results, file =paste0(output_dir,"EH22_dds_diatom_clusters.rds"))
print(Sys.time() - start_time)
```
```{r}
dds.results<- readRDS("/Users/houlin/Desktop/Research/Ecohab/Ecohab22/metatranscriptoms/DESeq2_results/EH22_dds_diatom_clusters.rds")
```

```{r}
treatment_levels <- levels(dds.results$Diatom_clusters)
output_dir <- "/Users/houlin/Desktop/Research/Ecohab/Ecohab22/metatranscriptoms/DESeq2_results/"
# Loop through each pair of treatments 
results_list <- list()
LFC_results_list<-list()
for (i in 1:(length(treatment_levels) - 1)){
  for (j in (i + 1):length(treatment_levels)) {
    contrast <- c("Diatom_clusters", treatment_levels[i], treatment_levels[j])
    res <- results(dds.results, contrast = contrast)
    # shrinkage to remove extreme values
    resLFC <- lfcShrink(dds.results,contrast = contrast, type="normal")
    comparison_name <- paste(treatment_levels[i], "_vs_", treatment_levels[j], sep = "")
    par(mfrow=c(1,2), mar=c(4,4,2,1))
    plotMA(res, main="raw")
    plotMA(resLFC, main="apelgm")
    print(comparison_name)
    mtext(comparison_name, side=3, line=1, outer=TRUE, cex=1.5)
    results_list[[comparison_name]] <- res
    LFC_results_list[[comparison_name]]<-resLFC
    
    resSig_filter<- resSig_filter[order(-resSig_filter$log2FoldChange), ]
    write.csv(resSig_filter, file = paste0(output_dir,"Diatom_DESeq2_genecounts_", comparison_name, ".csv"))
    
    resSig_filter<- resSig_filter[order(-resSig_filter$log2FoldChange), ]
    write.csv(resSig_filter, file = paste0(output_dir,"Diatom_DESeq2_genecounts_", comparison_name, "_LFC.csv"))
  }
}
```
```{r}
treatment_levels <- levels(dds.results$Diatom_clusters)
comparison_name <- paste(treatment_levels[2], "_vs_", treatment_levels[3], sep = "")
print(comparison_name)
contrast <- c("Diatom_clusters", treatment_levels[2], treatment_levels[3])# 1,2,3 , nonbloom, bloom-1,bloom-2
res <- results(dds.results, contrast = contrast)
resLFC <- lfcShrink(dds.results,contrast = contrast, type="normal")
par(mfrow=c(1,2), mar=c(4,4,2,1))
plotMA(res, main="Bloom-1 vs Bloom-2 raw")
plotMA(resLFC, main="Bloom-1 vs Bloom-2 LFC")
```


```{r}
resLFC_annot <- as.data.frame(res)# significant only if padj<0.05 and not  na
resLFC_annot$geneid <- rownames(resLFC_annot)
annotation$geneid<-rownames(annotation)# Merge with your annotation file
resLFC_annot <- inner_join(resLFC_annot, annotation, by = "geneid")
rownames(resLFC_annot)<-resLFC_annot$geneid
```

```{r, fig.width=5, fig.height=5}
#resLFC_annot$flag <- (resLFC_annot$padj < 0.1) & resLFC_annot$ISIP_gene != "" 
#unique(resLFC_annot[grepl("silicon", resLFC_annot$MCL_anno, ignore.case = TRUE),]$MCL_anno)
resLFC_annot$flag <- (resLFC_annot$padj < 0.1) & grepl("flavodoxin", resLFC_annot$MCL_anno, ignore.case = TRUE)
png("your_plot_name.png", width = 1800, height = 1600, res = 300)  # dimensions in pixels, res = resolution (dpi)
plotMA(res, main = "Bloom-1 vs Bloom-2 Flavodoxin",
       cex.lab = 1,    # Larger axis label size
       cex.axis = 0.8,   # Larger tick mark size
       cex.main = 1,
       mgp = c(2, 0.6, 0))
# Circle ISIP genes on top
with(resLFC_annot[resLFC_annot$flag, ], {
  points(baseMean, log2FoldChange,
         pch = 21, 
         bg = NA, 
         col = "red",  # red circle
         cex = 0.8, 
         lwd = 1.5)
})

dev.off()
```
```{r, fig.width=5, fig.height=5}
plotMA(res, main = "Diatom Bloom-1 vs Bloom-2 ",
       cex.lab = 1,    # Larger axis label size
       cex.axis = 0.8,   # Larger tick mark size
       cex.main = 1,
       mgp = c(2, 0.6, 0))

# Add points for ISIP genes
ptsize=0.6
with(resLFC_annot[(resLFC_annot$padj < 0.1) & resLFC_annot$ISIP_gene != "", ], {
  points(baseMean, log2FoldChange, col = "black", pch = 16,cex=ptsize)})

# Add points for N_genes
with(resLFC_annot[(resLFC_annot$padj < 0.1)& grepl("flavodoxin", resLFC_annot$MCL_anno, ignore.case = TRUE), ], {
  points(baseMean, log2FoldChange, col = "orange", pch = 17,cex=ptsize)})

with(resLFC_annot[resLFC_annot$padj < 0.1 & grepl("plastocyanin", resLFC_annot$MCL_anno, ignore.case = TRUE), ],
     {points(baseMean, log2FoldChange, col = "green", pch = 18,cex=ptsize)})

with(resLFC_annot[resLFC_annot$padj < 0.1 &
      grepl("high-affinity iron transporter", resLFC_annot$MCL_anno, ignore.case = TRUE), ],
     points(baseMean, log2FoldChange, col = "red", pch = 18,cex=ptsize))
legend("topleft",   # <- Key part!
       legend=c("ISIP", "Flavodoxin", "plastocyanin","Iron Transporter"),
       col=c("black", "orange", "green","red"),
       pch=c(16,17,18),cex=0.6,
       title="Gene Types") 
```

######## change the MCL_anno by making sure it matches ISIP_gene #########
```{r}
resLFC_annot$MCL_anno <- ifelse(!is.na(resLFC_annot$ISIP_gene) & resLFC_annot$ISIP_gene != "",
                                resLFC_annot$ISIP_gene,
                                resLFC_annot$MCL_anno)

# Then N_genes: overwrite MCL_anno if N_genes is not NA or empty
resLFC_annot$MCL_anno <- ifelse(!is.na(resLFC_annot$N_genes) & resLFC_annot$N_genes != "",
                                resLFC_annot$N_genes,resLFC_annot$MCL_anno)

resLFC_annot$MCL_anno <- ifelse(!is.na(resLFC_annot$dab_gene) & resLFC_annot$dab_gene != "",
                                resLFC_annot$dab_gene,resLFC_annot$MCL_anno)

resLFC_annot$MCL_anno <- ifelse(!is.na(resLFC_annot$SiLA_gene),resLFC_annot$SiLA_gene,
                                resLFC_annot$MCL_anno)

resLFC_annot$MCL_anno <- ifelse(grepl("flavodoxin", resLFC_annot$MCL_anno, ignore.case = TRUE),
                                "Flavodoxin",resLFC_annot$MCL_anno)

resLFC_annot$MCL_anno <- ifelse(grepl("glutamine synthetase", resLFC_annot$MCL_anno, ignore.case = FALSE),
                                "GSIIA",resLFC_annot$MCL_anno)

resLFC_annot$MCL_anno <- ifelse(grepl("NirB|nitrite reductase \\[NAD\\(P\\)H\\], small subunit", 
        resLFC_annot$MCL_anno, ignore.case = FALSE),
  "NADH-Nir",resLFC_annot$MCL_anno)

resLFC_annot$MCL_anno <- ifelse(grepl("ferredoxin-nitrite reductase|NirG", resLFC_annot$MCL_anno, ignore.case = FALSE),
  "Fd-Nir",resLFC_annot$MCL_anno)

resLFC_annot$MCL_anno <- ifelse(grepl("GLUTAMINE SYNTHETASE", resLFC_annot$MCL_anno, ignore.case = FALSE),
                                "GSIII",resLFC_annot$MCL_anno)
resLFC_annot$MCL_anno <- ifelse(grepl("glutamate synthase", resLFC_annot$MCL_anno, ignore.case = FALSE),
                                "GOGAT",resLFC_annot$MCL_anno)
```


```{r}
FC_mean <- resLFC_annot %>%
  group_by(MCL_anno) %>%
  summarise(mean_log2FC = mean(log2FoldChange, na.rm = TRUE))

# Weighted mean
FC_weighted_mean <-resLFC_annot %>%
  group_by(MCL_anno) %>%
  summarise(weighted_log2FC = weighted.mean(log2FoldChange, baseMean, na.rm = TRUE))
```


```{r}
rownames(resLFC_annot)<-resLFC_annot$geneid
output_dir <- "/Users/houlin/Desktop/Research/Ecohab/Ecohab22/metatranscriptoms/DESeq2_results/"
write.csv(resLFC_annot, file = paste0(output_dir,"Diatom_DESeq2_genecounts_bloom-1_bloom-2_anno.csv"))
```

